Experience Report: Haskell in Computational Biology

### 2012-03-12 mrfy_experience_report.tex
complete edited draft

In MRFy, the \texttt{score} function calls \texttt{viterbi} several
 times
for each placement in each search generation.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
We~found them easy to parallelize
using
\texttt{Control.Parallel},
simply substituting
\texttt{parmap rseq} for \texttt{map}.
Even when there is just one placement per generation,
MRFy can keep six cores busy,
and when our search algorithm uses multiple placements per generation,
it~can use all forty-eight cores on a compute server.

In~\secref{viterbi} above, the example case for the \texttt{viterbi}
function computes not just a score but also an optimal path.
MRFy's \texttt{search} function, in \secref{search}, uses only the
score.
Even though Haskell evaluation is lazy, the \texttt{viterbi} function
is still allocating thunks.
We~tried cloning and modifying the \texttt{viterbi} code to compute
only a score, with no path, and this change improved 
run-time performance by nearly 50\%.
But when \texttt{search} finally decides on a placement, we need a
version of \texttt{viterbi} that \emph{does} compute path, so we can
extract the optimal parse for that placement.

How could we keep the performance improvement without having to
maintain two versions of \texttt{viterbi}?
In~Lisp or Ruby we would have used macros or metaprogramming,
but we were not confident of our ability to metaprogram a solution
using Template Haskell.
Instead, we replaced the primitive cons in \texttt{viterbi} with a
call to a function of the same type, which we passed~in.
Passing the primitive~``\texttt{:}'' gives us a path,
and passing \verb+\_ _ -> []+ gives us the same
speedup as the cloned and modified~code.
But~although
this trick is simple and easy to implement, 
we~have an uneasy feeling that it may work only because of undocumented
properties of GHC's inliner, which may be subject to change without
notice.

@@

In two major areas, our lack of skills or knowledge resulted in ugly
Haskell.
The first is profiling.
GHC's profiling is based on cost centers \cite{sansom-pj},
and although GHC 
will happily add cost centers to top-level 
functions, much of our code is in nested functions.
We~know of no better technique than to add cost-center annotations to
each guard clause and nested function,
but these annotations made our code so ugly that we felt compelled to
remove them as soon as possible.
And without the annotations, we found GHC's call-site and allocation
profiles did not provide enough information to be useful.
We~miss valgrind, whose callgrind profiler gives us enough information
without \emph{any} annotations in the source.

Our other ugly code came from  attempts to debug
run-time errors.
MRFy uses a Markov random field which is represented in a legacy file
format that uses a poorly documented representation---when beta
strands overlap and are doubly paired, the invariants of the
representation are not clear.
Dealing with this representation is difficult no matter what
programming language is used, and faults are to be expected.
Using GHC, we had a hard time diagnosing run-time errors.
Extensive use of 
\texttt{trace} 
littered our code,
even when moved from our main code to wrapper functions.
We~weren't aware of the backtrace feature of GHC's profiler,
and even after we discovered that feature, the results weren't good enough.
Specifically, because the backtrace feature required profiling mode,
it could only be used in very small test cases that didn't always trigger
run-time errors, and one of us was unable to overcome difficulties in building
various libraries in profiling mode.
In~August 2011, 
Lennart Augustsson said that the biggest advantage of Strict
Haskell is getting a stack trace on error,
and Simon Marlow said that he may have figured out how to track call
stacks properly in a lazy functional language.
We~can't wait.

@@

When performance has mattered, the members of our group, like other
computational biologists, have used~C++.
We~compare our experience with three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tool or library it uses is \texttt{zlib}.
Its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries.
Its initial development took a year and a half.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands.
It~comprises about 1,200 lines of~Haskell
It~uses several external tools and libraries, of which the most
notable are PADS-Haskell, the \path{Data.MemoCombinators} library,
the BioHaskell bioinformatics library, and the \path{Data.Vector} library.
Its initial development took about three months.
\remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}
Like much research software, all three of these tools were written
 in~haste.
We~have experience modifying the older tools.

We~modified Matt to use information about sequences as well as structure.
The modification added 2,000 lines of code, and it calls
external sequence aligners that we did not write.
We~thought the modification would take three months, 
but it took most of a~year.
Matt uses such
data structures such as mutable oct-trees, vectors, and arrays.
It~uses clever pointer arithmetic.
The mutable data structures were difficult to 
repurpose, and the pointer arithmetic was too clever for its own good: 
nearly every change resulted in new segfaults.

We~had hoped to extend Matt further, with support for partial alignments,
which we~expected to require only 
a cosmetic manipulation of the 
output, but
this feature wound up
 requiring deep information about some data structures,
and we had to give~up.
We~believe we could write an equivalent tool in~Haskell,
with most of Matt's performance, in at most nine months.

@@

On the protein structures for which it was originally designed, SMURF can 
compute the parse of a query sequence to a MRF based on those structures in 
less than 10 seconds on a typical workstation.
However, on somewhat more 
complex protein structures, SMURF may take two to three minutes on the same 
hardware. 
On the most complex protein structures, SMURF exhausts all memory on 
any system we have available, and must be killed. 
In contrast, MRFy's runtime 
for any of these protein structures is typically less than 30 seconds. 
This 
meets the performance goals we had, and means that MRFy is fast enough to scale 
to whole-genome scans.

@@

@@

If~you are a computational biologist, 
you can be productive in Haskell without extensive
experience.
Two~of us (Daniels and Gallant) are graduate students.
Daniels has taken a graduate seminar in functional programming, which
included some Haskell;
Gallant has taken a programming-languages course which included
significant functional programming but no Haskell.
Ramsey is a professor who has used Haskell and other functional
languages for several medium-sized projects,
but his contribution was limited to minor ranting and refactoring.

\subsection{Barriers}

@@

We~faced another barrier in testing.
One of us (Ramsey) kept ranting that we should test our code using 
QuickCheck \cite{claessen:quickcheck},
but we could never figure out~how.
We~were stumped by two problems:
how to break down a complex module into parts that can usefully be
unit tested,
and how to implement QuickCheck generators for representations of complex
data like beta-strand placements.
We~know that QuickCheck is a powerful and proven tool,
yet we were not smart enough to figure out how to use~it.

Like other functional programmers, we~have found that often enough,
once we have the types right, the code is right.
But~much of our work involves computing with arrays and array indices,
and in that domain, types don't help much.
And we haven't been able to identify any systematic way of debugging
run-time errors in Haskell programs.
GHC's profiler can provide stack traces, but we~found this information
difficult to discover, and as noted above,
there are barriers to profiling.
We're~aware that debugging lazy functional programs has been a topic
of some research,\remark{Do we cite Hood, Hat, or Freya?}
but one of the biggest barriers we encountered to using Haskell is
that we have had to abandon our old approaches to debugging,
and we~see no clear path to follow.

@@

If~you want to use Haskell in your research, we~believe that you must
have enough experience with functional programming that you can build
\emph{all} the code you need, not only the code that is easy to write
in a functional language.
Implementing Viterbi's equations in Haskell was pure joy.
Writing an iterative search in purely functional style was not
difficult.
When we had to transform data in the HMMER file format, \emph{without}
using mutable state the way the C++ code does, we~struggled.

@@

A~little knowledge of and a lot of love for functional programming was
enough for us to carry out a successful research project in a language
that computational biologists seldom use.
If~you \emph{want} to use Haskell---or one of your graduate students
wants to use Haskell---you can
succeed.

>>>>

MRFy calls \verb+vee'+ several
 times
for each placement in each search generation.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
We~found them easy to parallelize
using
\texttt{Control.Parallel}:
we substituted
\texttt{parmap rseq} for \texttt{map}.
Even when there is just one placement per generation,
MRFy can keep six cores busy.
Given multiple placements per generation,
it~can use forty-eight cores.

In~\secref{viterbi} above, the example case for the \verb+vee'+
function computes not just a score but also an optimal path.
MRFy's \texttt{search} function, in \secref{search}, uses only the
score.
Even though Haskell evaluation is lazy, the \verb+vee'+ 
still allocates thunks that could compute paths.
We~cloned \verb+vee'+ and modified it to compute
only a score, with no path, and this change improved 
run-time performance by nearly~50\%.
But when \texttt{search} finally decides on a placement, we need a
version of \verb+vee'+ that \emph{does} compute a path, so we can
extract the optimal parse for that placement.

How could we keep the performance improvement without having to
maintain two versions of \verb+vee'+?
In~Lisp or Ruby we would have used macros or metaprogramming,
but we were not confident of our ability to metaprogram a solution
using Template Haskell.
Instead, we replaced the primitive cons in \verb+vee'+ with a
call to a function of the same type, which we passed~in.
Passing the primitive~``\texttt{:}'' gives us a path,
and passing \verb+\_ _ -> []+ gives us the same
speedup as the cloned and modified~code.
But~although
this trick is simple and easy to implement, 
we~have an uneasy feeling that it may work only because of undocumented
properties of GHC's inliner, which may be subject to change without
notice.

@@

We~think \verb+vee'+ is beautiful, but we wrote some ugly code, too,
because we lack skills and knowledge in two areas.
The first is profiling.
GHC's profiling is based on cost centers \cite{sansom-pj},
and although GHC 
will happily add cost centers to top-level 
functions, much of our code is in nested functions.
We~know of no better technique than to add cost-center annotations to
each guard clause and nested function,
but these annotations made our code so ugly that we felt compelled to
remove them as soon as possible.
And without the annotations, GHC's call-site and allocation
profiles did not give us enough information to be useful.
We~miss valgrind, whose callgrind profiler gives useful information
without \emph{any} annotations in the source.

Our other ugly code came from  attempts to debug
run-time errors.
Our~group's legacy file
format is poorly documented: when beta
strands overlap and are doubly paired, the invariants of the
representation are not clear.
These files are hard to deal with no matter what
programming language is used, and faults are to be expected.
But using Haskell, we had a hard time diagnosing run-time errors.
Calls to
\texttt{trace} 
littered our code,
even when relegated to wrapper functions.
We~weren't aware of the backtrace feature of GHC's profiler,
and even after we discovered that feature, the results weren't good enough:
the profiler can be used only on very small test cases that didn't always trigger
run-time errors.
In~August 2011, 
Lennart Augustsson said that the biggest advantage of Strict
Haskell is getting a stack trace on error,
and Simon Marlow said that he may have figured out how to track call
stacks properly in a lazy functional language.
We~can't wait.

@@

When performance has mattered, members of our group, like other
computational biologists, have used~C++.
We~compare our experience with three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tool or library it uses is \texttt{zlib}, and
its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries, and
its initial development took a year and a half.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands.
It~comprises about 1,200 lines of~Haskell
It~uses several external tools and libraries, of which the most
notable are PADS-Haskell, the \path{Data.MemoCombinators} library,
the BioHaskell bioinformatics library, and the \path{Data.Vector} library.
Its initial development took about three months.
%%%% \remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}
Like much research software, all three of these tools were written
 in~haste.
We~have experience modifying the older tools.

We~modified Matt to use information about sequences as well as structure.
The modification added 2,000 lines of code, and it calls
external sequence aligners that we did not write.
We~thought the modification would take three months, 
but it took most of a~year.
Matt uses such
data structures such as mutable oct-trees, vectors, and arrays.
It~uses clever pointer arithmetic.
The mutable data structures were difficult to 
repurpose, and the pointer arithmetic was \emph{too} clever: 
nearly every change resulted in new segfaults.

We~had hoped to extend Matt further, with support for partial alignments,
which we~expected to require only 
a cosmetic manipulation of the 
output.
But
this feature wound up
requiring deep information about Matt's data structures,
and we had to give~up.
We~believe we could write an equivalent tool in~Haskell,
with most of Matt's performance, in at most nine months.

@@

@@

We~can also make a preliminary comparison of performance.
SMURF can parse a simple protein structure, optimally, in under 10~seconds.
It~can parse some more complex structures in 2--3~minutes.
But when asked to parse the most complex structures, it runs out of
memory and has to be killed, even on a system with 128GB of~RAM.
By~contrast, MRFy can parse any of these structures, approximately,
in 30~seconds.
This performance
meets our goals, and it means that MRFy can be used to scan whole genomes.

@@

If~you are a computational biologist, 
you can be productive in Haskell, and you don't need extensive
experience.
Two~of us (Daniels and Gallant) are graduate students.
Daniels has taken a graduate seminar in functional programming, which
included some Haskell;
Gallant has taken a programming-languages course which included
significant functional programming but no Haskell.
Ramsey is a professor who has used Haskell 
for medium-sized projects,
but his contribution to MRFy was limited to minor ranting and refactoring.

\subsection{Barriers to be overcome}

@@

We~faced another barrier in testing.
One of us (Ramsey) kept ranting that we should test our code using 
QuickCheck \cite{claessen:quickcheck},
but we could never figure out~how.
We~were stumped by two problems:
how to break down MRFy into parts that could be
unit-tested,
and how to implement QuickCheck generators for representations of complex
data like beta-strand placements.
We~know that QuickCheck is a powerful and proven tool,
but we could figure out how to exploit~it.

Like other functional programmers, we~have found that 
once we have our types right, our code is often right.
But~MRFy computes with arrays and array indices,
and in that domain, types don't help much.
Bounds violations lead to run-time errors, which we have not been able
to identify any systematic way to~debug.
GHC's profiler can provide stack traces, but we~found this information
difficult to discover, and as noted above,
there are barriers to profiling.
We're~aware that debugging lazy functional programs has been a topic
of some research,\remark{Do we cite Hood, Hat, or Freya?}
but one of the biggest barriers we encountered to using Haskell is
that we have had to abandon our old approaches to debugging,
and we~see no clear path to follow.

@@

If~you want to use Haskell in your research, we~believe that you must
have enough experience with functional programming that you can build
\emph{all} the code you need, not only the code that is easy to write
in a functional language.
Implementing Viterbi's equations in Haskell was pure joy.
Writing an iterative search in purely functional style was~easy.
Transforming data in the HMMER file format, \emph{without}
using mutable state the way the C++ code does, was difficult.

@@

A~little knowledge of and a lot of love for functional programming
enabled us to carry out a successful research project in a language
that computational biologists seldom use.
If~you \emph{want} to use Haskell---or one of your graduate students
wants to use Haskell---you can
succeed.
